---
title: "Chick Permutation"
author: "Daniel Coker"
format: html
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(ggplot2)
library(infer)
data("ChickWeight")
```

1.  Chick Weight Dataset

The "Chick Weight" dataset records the weight of chicks measured throughout time. It is one of the include datasets in R. The variables included are weight, time, chick (unique idenitier of each chick), and diet (coded as 1-4) where the higher number represents the percentage of protein in a diet plan with 1 representing a diet that contains 10% protein and a 4 representing a diet that contains 40% protein. 

My research question for my premuatation that I want to explore is is there a significant difference between weight gain for chicks between a low protein diet, 1, and a high protein diet, 4? Thus, for this project, I only will be using the 'weight' and 'diet' variables. My $H_0$ would mean that there is no relationship between diet and weight while my $H_A$ suggest that there is a relationship between them. 

Here is the permutation test:
```{r}
chick_data <- ChickWeight |> 
  filter(Time == max(Time)) 

#Keep only Diet 1 & 4
chick_data_filtered <- chick_data |> 
  filter(Diet %in% c(1, 4))

#Function to compute difference in means
compute_obs_diff <- function(data, diet1, diet2) {
  data |> 
    filter(Diet %in% c(diet1, diet2)) |> 
    group_by(Diet) |> 
    summarize(mean_weight = mean(weight), .groups = "drop") |> 
    summarize(diff = diff(mean_weight)) |> 
    pull(diff)
}

#Compute observed difference in means
obs_diff_value <- compute_obs_diff(chick_data_filtered, diet1 = 1, diet2 = 4)

#Permutation tet
set.seed(47)
perm_results <- chick_data_filtered |> 
  specify(weight ~ Diet) |> 
  hypothesize(null = "independence") |> 
  generate(reps = 4700, type = "permute") |> 
  calculate(stat = "diff in means", order = c("1", "4"))

p_value <- mean(map_dbl(perm_results$stat, ~ .x >= obs_diff_value))

#Plot null distribution
ggplot(perm_results, aes(x = stat)) +
  geom_histogram(binwidth = 5, fill = "lightgreen", color = "black") +
  geom_vline(xintercept = obs_diff_value, color = "red", linetype = "dashed") +
  ggtitle("Null Distribution") +
  xlab("Difference in Mean Weight") +
  ylab("Frequency")

p_value  
```
For my project, I first filtered out all the other data I was not using, only keeping my chick's weights and diets 1 and 4. Then, I computed the observed difference from the mean of weight between diet 1 and 4. Using this, I then generate 4,700 permutation samples all using a different premutation for my test. From this, I then took the p-value of the premutation tests, giving me a low p-value of ~.0066 (sidenote: the p-value would have gone lower if I increased the repititions but chirp chirp). 

The null distribution histogram visualizes the differences in mean weight between diet 1 and 4 under the $H_0$ assuming it is true. Each bar represents the frequency of a particular difference in means across our 4,700 permutation samples. The red dashed line represents the observed difference in means from the actual dataset, showing how extreme having a result that fulfills the $H_0$ is.

It is clear that the p-value of ~.0066 suggest that diet and a chick's weight are significantly related. Since our observed difference line (dashed red line) is near the far-right tail, there is a very unlikely chance that this would occur randomly, thus, we can disregard the $H_0$. There is a clear relationship between a high-protien diet and weight.   
