---
title: "Whatcha Tweeting Mr. President? Analyzing the Tweets of President Barack Obama"
subtitle: "Final Presentation – May 13, 2025"
author: "Daniel Coker"
format: revealjs
---

# Introduction

- **Project Overview:** Analysis of Barack Obama's Tweets during his presidency (2009-2017).
- **Dataset:** National Archives collection of Obama's Tweets.
- **Objectives:**
  - Explore tweeting patterns over time.
  - Analyze tweet frequency and sentiment.
  - Investigate most frequently used words and phrases.

# Background

- **Context:**
  - Barack Obama was the first US President to actively use social media, specifically Twitter, throughout his presidency and campaign.
  - The tweets that were analyzed were collected from Barack Obama's personal Twitter account. No other affilated tweets were collected.
- **Data Source:**
  - National Archives' Barack Obama Presidential Library.


# Tweet Frequency

- **Monthly Tweet Analysis:**
- Significant peaks were observed through the latter half of his presidency especially between the summers of  2015 & 2016.
- Potential causes include political campaigns, Black Lives Matter protests, and major policy announcements could help explain certain spikes.

```{r}
library(tidyverse)
library(tidytext)
library(ggplot2) 


tweets <- read_csv("tweets.csv")
tweets <- tweets |>
  mutate(Date = as.Date(timestamp), MonthYear = floor_date(Date, "month")) 

tweet_counts <- tweets |>
  count(MonthYear)

# Plotting
ggplot(tweet_counts, aes(x = MonthYear, y = n)) +
  geom_col(fill = "navy", alpha = 0.7) +
  theme_light() +
  labs(title = "Frequency of Obama's Tweets by Month",
       x = "Month-Year",
       y = "Number of Tweets") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



# Heatmap Analysis (Mentions of 'America')

- **Focus:** Tweets that specifically mention "America".
-Meant to identify periods of high patriotic or policy-driven content.

```{r}
tweets |>
  filter(str_detect(text, "(?i)\\bamerica\\b")) |>  
  mutate(Year = year(Date), Month = month(Date, label = TRUE)) |>
  count(Year, Month) |>
  ggplot(aes(x = Month, y = as.factor(Year), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "orange", high = "red") +
  theme_minimal() +
  labs(title = "Obama's Tweet Frequency Heatmap (Mentions of 'America')",
       x = "Month",
       y = "Year",
       fill = "Tweets")
```



# Word Frequency Analysis

- **Most Used Words:** I wanted to analyze Obama's most common tweeted words. 
-Wanted to see if there were any nuanced meanings/themes behind his diction through his time in office
- **Common Words I Excluded:** "you," "the," "to," "and," "a," etc.

```{r}
exclude_words <- c("you", "the", "to", "and", "a", "for", "on", "of", "rt")

word_counts <- tweets |>
  mutate(text = str_to_lower(text), text = str_replace_all(text, "[^a-z\\s]", "")) |>
  unnest_tokens(word, text) |>
  filter(!word %in% exclude_words) |>
  count(word, sort = TRUE) |>
  filter(n > 20)

# Plotting
word_counts |>
  top_n(20, n) |>
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "purple") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 20 Most Used Words in Obama's Tweets",
       x = "Word",
       y = "Frequency")
```



# Insights & Conclusions

- **Findings:**
  - It seems Obama used Twitter to communicate unity and national pride.
  - High activity observed in politically charged periods especially during the 2016 election process.
- **Ways to Improve:**
  - Explore further sentiment analysis of tweets.
  -Try to analyze tweets that predate this dataset
 -Connect voter turnout or approval with endorsed tweets



# That's All Folks!

- **Any questions?**
